---
title: "Practical Machine Learning - Week 4"
author: "PGM"
date: "18/11/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())
setwd('/Users/PGM-MAC/Desktop/Johns Hopkins')
```

### Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

### Cleaning data

```{r}
## Download the CSV files with the data
fileURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(fileURL, destfile = "./pml-training.csv", method = "curl")
dateDownloaded <- date()

fileURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(fileURL, destfile = "./pml-testing.csv", method = "curl")
dateDownloaded <- date()

## Loading the data from the CSV files downloaded
pml_training <- read.csv("./pml-training.csv", sep = ",", header = TRUE)
head(pml_training)

pml_testing <- read.csv("./pml-testing.csv", sep = ",", header = TRUE)
head(pml_testing)
```

Search for the strings regarding the accelerometers on 4 locations (belt, forearm, arm, dumbbell)

```{r}
choices <- c("accel_belt", "accel_forearm", "accel_arm", "accel_dumbbell")
columns <- grepl(paste(choices, collapse = "|"), colnames(pml_training))
```

Manually making that the first and last elements true to select those extra columns:

```{r}
# Extracting the data in those columns only
specificTraining <- pml_training[, columns]

## Removing choices and columns
rm(choices, columns)
```

Looking briefly at the resulting data:

```{r}
summary(specificTraining)
```

The amount of NA's is so high on the variables 'var_accel' & 'var_total_accel' from the 4 locations that it those columns will be removed from the pool of independent variables.

```{r}
choices <- c("var_accel", "var_total_accel")
columns <- grepl(paste(choices, collapse = "|"), colnames(specificTraining))

# Extracting the data in those columns only
specificTraining <- specificTraining[, !columns]

## Removing choices and columns
rm(choices, columns)
```

Adding the "classe" variable to the specificTraining dataset:

```{r}
# Adding classe
specificTraining <- cbind(specificTraining, classe = pml_training$classe)
```

### Data partition to evaluate out-of-sample errors

As the model cannot be tested with pml_testing without knowing how it performs, it is necessary to split the training data in 2 groups:

```{r}
library(caret)
set.seed(28322)

# Generating a possible partition (80% / 20%)
inTrain <- createDataPartition(y = specificTraining$classe, p = 0.80, list = FALSE)

# Assigning most of the data to the subtraining group and the rest to the subtesting group 
subtraining  <- specificTraining[inTrain, ]
subtesting  <- specificTraining[-inTrain, ]
```

### Model development

Building a random forest model:

```{r}
library(randomForest)

set.seed(28322)

# Building a random forest model
modFit <- randomForest(classe ~ ., data = subtraining)
modFit
```

Showing the results visually for the iterations made during the model creation:

```{r}
plot(modFit)
```

Dotchart of variable importance as measured by the random forest model just created:

```{r}
varImpPlot(modFit)
```

### Application of model to subtesting dataset

Predicting the classe values for subtesting:

```{r}
# A variable classe is added to the testing dataset with the predictions
prediction <- predict(modFit, newdata = subtesting)

# Creating the confusion matrix for the results achieved
confusion <- confusionMatrix(prediction, subtesting$classe)
confusion
```

Accuracy of 94.83%, so that seems interesting.

From this result, it is possible to have a reference of the performance from the calculated model:

```{r}
out_of_sample_error <- 1 - confusion$overall[1]
names(out_of_sample_error) <- "Error"
out_of_sample_error
```

### Application of model to global testing dataset

Predicting the classe values for pml_testing:

```{r}
# A variable classe is added to the testing dataset with the predictions
pml_testing$classe <- predict(modFit, newdata = pml_testing)

# The predictions to be provider to the automatic checker
pml_testing$classe
```